Kepler's Demographic Cliff: Ages of Young Stars and Planets in the Kepler Field

### 2024-05-09 paper/ms.tex
update ms

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.48\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
		}
		\subfloat{
			\includegraphics[width=0.48\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
		{\bf Rotation periods for Kepler target stars (left) and known
			planet hosts (right)}.  
		The gray lines are ``mean fits'' 
		to the slow rotation sequences of open
		clusters.  The stellar sample (left) includes only apparently single stars
		near the main sequence with $\log g$$>$4.2, RUWE$<$1.4,
		and 3800$<$$T_{\rm eff}$/K$<$6200.
		The color map has a linear stretch.  %between one and ten stars per bin.
		The planet-hosts (right) require
		the same stellar cuts, and include only the confirmed and candidate planets
		described in Section~\ref{subsec:plflags}.
		An order of magnitude estimate for the number of young planets discovered
		by Kepler follows by counting planets below the mean rotational isochrones.
	}
	\label{fig:prot_vs_teff}
\end{figure*}

@@

\begin{figure}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.47\textwidth]{li_vs_teff_koi_X_JUMP_eagles_logy.pdf}
	\end{center}
	\vspace{-0.6cm}
	\caption{
		{\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
			for planet-hosting stars.} These measurements were made from
		Keck/HIRES spectra collected between 2009 and 2024 that comprised
		\nhireshours\,hours of open-shutter time.  Lines are the ``mean''
		isochrones from the same models as in Figure~\ref{fig:models}; the
		intrinsic dispersion beyond $\approx$1\,Gyr becomes much larger
		than changes in the mean.  The logarithmic scale highlights the
		dynamic range of the detections and upper limits.  Some stars with
		lithium detections do not have detected rotation periods, and
		vice-versa.
		\label{fig:li_vs_teff}
	}
\end{figure}

@@

\subsection{Stellar Quality Flags}
\label{subsec:flags}
We calculated gyrochrone age posteriors for all
\nuniqstarsantosrotteffcut\ stars with reported rotation periods and
effective temperatures between 3800\,K and 6200\,K.  To flag stars for
which we suspect these ages may not in fact be valid, we built a set
of quality flags, which we condensed into a single binary number,
$Q_{\rm star}$.  When and how the flags represented in this should be
applied depends on the question being asked.  If the goal is to
construct a false-positive free sample, they could all be applied.  If
the goal is to construct a complete sample, then consider the examples
of Kepler~1627Ab ($\approx$40\,Myr) and Kepler~51c
($\approx$625\,Myr).  The former has a high RUWE due to a resolved
binary companion \citep{Bouma_2022a}; the latter is on a grazing orbit
\citep{2014ApJ...783...53M}.  We leave selection for or against such
cases as a decision to the user.

@@

The rotation periods for the star and planet samples are shown
in Figure~\ref{fig:prot_vs_teff}.  Out of the \nkeplerstars\ stars
observed by Kepler, requiring reported rotation periods and a reliable
DR3 crossmatch yields \nuniqstarsantosrot\ stars.  Further requiring
these stars to have temperatures 3800$<$$T_{\rm
eff}$/K$<$6200, yields \nuniqstarsantosrotteffcut\
stars.  Further requiring the stars to be apparently single, near the main
sequence, with $\log g$$>$4.2, $M_{\rm G}>3.9$, and not flagged as
eclipsing binaries yielded \nuniqstarsantosallbutruwe\ rotators
potentially amenable for gyrochronology.  Finally, imposing RUWE$<$1.4
and requiring no possible blends ($\Delta G$<2.5) within 4$''$ leaves
\nuniqstarsantosrotgyroappl\ stars for which gyrochronology is likely
to be valid.

@@

	\vspace{-0.35cm}
    \subfloat{
    \includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000_preciseagesonly.pdf}
    }
  \end{center}
  \vspace{-0.6cm}
  \caption{{\bf Kepler's demographic cliff,} visible in the rotation-derived
    ages of stars (left) and planet hosts (middle) in the Kepler field.
    The top row corresponds to all
    stars with $T_{\rm eff}\in[3800,6200]\,$K for which we calculated rotation-based ages.
    The darker lines impose quality cuts
    on factors such as binarity, crowding, and the star's evolutionary
    state.
    The bottom row selects only stars with $T_{\rm eff}\in[4400,5400]\,$K, which
    have relatively precise ages near 2\,Gyr due to their fast spin-down
    \citep[see][]{Bouma_2023}.
    The statistical uncertainty for a mean individual star at 1, 2, and 3\,Gyr is shown, and is identical across each row.
    Finally, the right panel compares the rotation-based age distributions against star formation histories derived using CMD fitting.
    Completeness is near unity until $t$$\lesssim$3\,Gyr
    \citep{2022ApJ...937...94M}.
    \label{fig:hist_tgyro}
  }
\end{figure*}

@@

Our approach to answering these questions was to curate a sample of
stellar rotation periods, lithium equivalent widths, and temperatures
using archival and new data.  We derived new ages
using empirical interpolation-based methods that have been
calibrated against an extensive set of open cluster data.

>>>>

\begin{figure*}[!t]
	\begin{center}
		\subfloat{
			\includegraphics[width=0.48\textwidth]{prot_teff_Santos19_Santos21_dquality.pdf}
		}
		\subfloat{
			\includegraphics[width=0.48\textwidth]{koi_mean_prot_teff_koi_X_S19S21dquality_keepgrazing.pdf}
		}
	\end{center}
	\vspace{-0.5cm}
	\caption{
		{\bf Rotation periods for Kepler target stars (left) and known
			planet hosts (right)}.  
		The gray lines are ``mean fits'' 
		to the slow rotation sequences of open
		clusters.  The stellar sample (left) includes only apparently single stars
		near the main sequence with $\log g$$>$4.2, RUWE$<$1.4,
		and temperatures of 3800--6200\,K.
		The color map has a linear stretch.  %between one and ten stars per bin.
		The planet-hosts (right) require
		the same stellar cuts, and include only the confirmed and candidate planets
		described in Section~\ref{subsec:plflags}.
		An order of magnitude estimate for the number of young planets discovered
		by Kepler follows by counting planets below the mean rotational isochrones.
	}
	\label{fig:prot_vs_teff}
\end{figure*}

@@

\begin{figure}[!t]
	\begin{center}
		\leavevmode
		\includegraphics[width=0.47\textwidth]{li_vs_teff_koi_X_JUMP_eagles.pdf}
	\end{center}
	\vspace{-0.6cm}
	\caption{
		{\bf Equivalent widths (EW) of the \ion{Li}{1} 6708\,\AA\ doublet
			for planet-hosting stars.} These measurements were made from
		Keck/HIRES spectra collected between 2009 and 2024 that comprised
		\nhireshours\,hours of open-shutter time.  Lines are the ``mean''
		isochrones from the same models as in Figure~\ref{fig:models}; the
		intrinsic dispersion beyond $\approx$1\,Gyr becomes much larger
    than changes in the mean.  Some stars with lithium detections do
    not have detected rotation periods, and vice-versa.
		\label{fig:li_vs_teff}
	}
\end{figure}

@@

\subsection{Stellar Quality Flags}
\label{subsec:flags}
We calculated gyrochrone age posteriors for all
\nuniqstarsantosrotteffcut\ stars with reported rotation periods and
effective temperatures of 3800--6200\,K.  To flag stars for
which we suspect these ages may not in fact be valid, we built a set
of quality flags, which we condensed into a single binary number,
$Q_{\rm star}$.  When and how the flags represented in this should be
applied depends on the question being asked.  If the goal is to
construct a false-positive free sample, they could all be applied.  If
the goal is to construct a complete sample, then consider the examples
of Kepler~1627Ab ($\approx$40\,Myr) and Kepler~51c
($\approx$625\,Myr).  The former has a high RUWE due to a resolved
binary companion \citep{Bouma_2022a}; the latter is on a grazing orbit
\citep{2014ApJ...783...53M}.  We leave selection for or against such
cases as a decision to the user.

@@

The rotation periods for the star and planet samples are shown
in Figure~\ref{fig:prot_vs_teff}.  Out of the \nkeplerstars\ stars
observed by Kepler, requiring reported rotation periods and a reliable
DR3 crossmatch yields \nuniqstarsantosrot\ stars.  Further requiring
these stars to have temperatures of 3800--6200\,K, yields \nuniqstarsantosrotteffcut\
stars.  Further requiring the stars to be apparently single, near the main
sequence, with $\log g$$>$4.2, $M_{\rm G}>3.9$, and not flagged as
eclipsing binaries yielded \nuniqstarsantosallbutruwe\ rotators
potentially amenable for gyrochronology.  Finally, imposing RUWE$<$1.4
and requiring no possible blends ($\Delta G$<2.5) within 4$''$ leaves
\nuniqstarsantosrotgyroappl\ stars for which gyrochronology is likely
to be valid.

@@

	\vspace{-0.35cm}
    \subfloat{
    \includegraphics[width=.98\textwidth]{comp_hist_samples_koi_gyro_ages_hist_field_gyro_ages_20240430_maxage4000_preciseagesonly.pdf}
    }
  \end{center}
  \vspace{-0.6cm}
  \caption{{\bf Kepler's demographic cliff,} visible in the rotation-derived
    ages of stars (left) and planet hosts (middle) in the Kepler field.
    The top row corresponds to all
    stars with temperatures of 3800--6200\,K for which we calculated rotation-based ages.
    The darker lines impose quality cuts
    on factors such as binarity, crowding, and the star's evolutionary
    state.
    The bottom row selects only stars with temperatures of 4400-5400\,K, which
    have relatively precise ages near 2\,Gyr due to their fast spin-down
    \citep[see][]{Bouma_2023}.
    The statistical uncertainty for a mean individual star at 1, 2, and 3\,Gyr is shown, and is identical across each row.
    Finally, the right panel compares the rotation-based age distributions against star formation histories derived using CMD fitting.
    Completeness is near unity until $t$$\lesssim$3\,Gyr
    \citep{2022ApJ...937...94M}.
    \label{fig:hist_tgyro}
  }
\end{figure*}

@@

Our approach to answering these questions was to curate a sample of
stellar rotation periods, lithium equivalent widths, and temperatures
using archival and new data.  We derived new ages
using empirical methods that have been
calibrated against an extensive set of open cluster data.
